---
title: 'R: tips on structure, shortcuts and tidyverse'
output:
    html_document:
      toc: true
      toc_float: true
      theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# we want to print the error messages
knitr::opts_chunk$set(error=TRUE)
```


# Load libraries

```{r}
library(tidyverse)
```


# Learning resources

Note that in R, there are usually multiple ways to tackle the same problem. Subsetting, filtering, renaming, etc., can also be achieved via base R.  

* [R for data science (Wickham&Grolemund, 2017)](https://r4ds.had.co.nz/): a good guide to the tidyverse. Equally importantly, it can teach you about the phylosophy of tidy data, and the general "know-how" of data processing and analysis. The book has plenty of exercises for you to build "coding muscle". You can find unofficial solutions tot he exercises [here](https://jrnold.github.io/r4ds-exercise-solutions/).
* Cheatsheets covering tidyverse topics, which you can find [online](https://www.rstudio.com/resources/cheatsheets/), or in RStudio itself (Help > Cheatsheets).
* [Crawley (2013)](https://www.cs.upc.edu/~robert/teaching/estadistica/TheRBook.pdf): complete reference for basic coding, mathematics, statistics, etc.

# Workflow of the script

## Dataset: a modified mtcars

I will demonstrate most of the tidyverse functions using the `mtcars` dataset, containing 11 aspects of automobile design and performance for 32 automobiles. Type `?mtcars` if you want to learn more about this data frame.
```{r}
summary(mtcars)
```

Note that the `mtcars` dataset is a well-behaved, tidy dataset. In the real world, datasets are usually messy (sometimes through the fault of others, sometimes through our own). They might contain missing cases and/or duplicated observations and/or weird formats and/or whatever problem you can imagine.

Since untidy datasets are so prevalent, we will work with a more realistic dataset: I will add `NA`, `NaN` and `Inf`. In order to demonstrate column selection based on the type of information stored in the column, I add a column with characters (the original dataset contains only columns with numbers).

```{r}
# For simplicity and clarity, let's keep only a few columns, and a few rows
data <- mutate(mtcars, something = "a") %>%
  dplyr::select(mpg, cyl, disp, carb, something) %>% 
  dplyr::slice(1:10)
data[c(1:2, 5), 1] <- NA
data[c(2:3, 6), 2] <- NaN
data[4:6, 3] <- Inf
data
```

## The pipe `%>%`

I will demonstrate functions using pipes: `%>%`. The pipe `%>%` passes the object on the left to the first argument of the function on the right. The dot `.` can be used to specify the argument to which the object is passed. Check the `pull()` function lower in the script for an example of the use of the dot as placeholder.

Throughout the script, I will demonstrate the effect of the different functions by using the `structure()` function, or `str()` for short. Instead of writing: `str(data)`, I will use the pipe:

```{r}
data %>% 
  str()
```

Also note, the following functions will be demonstrated with a data frame, and there are additional formats for tabular data, such as tibble and data.table.


# Tidyverse: a few tips focusing on column operations


## Tidy selection of variables (`dplyr::select` + selection helpers)

The `dplyr::select()` takes a data frame and returns a data frame (or takes a tibble and returns a tibble, etc), allowing to select columns by name and/or numerical position. If you select a singe column, you will get a data frame with a single column, not a vector (see the `pull()` function below to get vectors from data frames). Note that the order in which variables are selected is preserved in the resulting data frame.

```{r}
data %>%
  dplyr::select(mpg) %>%
  str()

# use the minus sign to remove variables
data %>%
  dplyr::select(- mpg) %>%
  str()

# use c() to (de)select multiple columns
data %>%
  dplyr::select(-c(mpg, cyl)) %>%
  str()

# Use numerical position of columns. 1:3 is interpreted as range of consecutive values: from 1 to 3 including all integers in between
data %>%
  dplyr::select(1:3) %>%
  str()

# Remove columns base on position
data %>%
  dplyr::select(-c(1:3)) %>%
  str()

# Combine numbers and names
data %>%
  dplyr::select(1, disp) %>%
  str()

```

### Selection helpers

The use of `dplyr::select()` can be enhanced through selection helpers, of which you can see a non-exhaustive selection below. To learn more about selection of variables: `?dplyr_tidy_select`
Note that the default setting is `ignore.case = TRUE`, which means the default behavior is not case-sensitive.

```{r}
data %>%
  dplyr::select(everything()) %>%
  str()

data %>%
  dplyr::select(starts_with("c")) %>%
  str()

data %>%
  dplyr::select(ends_with("c")) %>%
  str()

data %>%
  dplyr::select(contains("c")) %>%
  str()

data %>%
  dplyr::select(contains("C")) %>%
  str()

# use the ignore.case argument to specify a case-sensitive behaviour
data %>%
  dplyr::select(contains("C", ignore.case = FALSE)) %>%
  str()
```

### Selection from a character vector (`all_of`, `any_of`)

In some cases, you might store the variables of interest in a character vector. The functions `all_of()` to `any_of()`, get the work done, but have slighlty different properties. The `all_of()` throws error if at least one element in the character vector is absent from the dataset


```{r}
my_selection <- c("mpg", "disp")
my_other_selection <- c("mpg", "gangman_style")

# all_of() for strict selection
data %>%
  dplyr::select(all_of(my_selection)) %>%
  str()

# use the minus sign to remove a selection
data %>%
  dplyr::select(-all_of(my_selection)) %>%
  str()

# we get an error message if the character vector contains names that are absent from the column names
data %>%
  dplyr::select(-all_of(my_other_selection)) %>%
  str()
```

On the other hand, `any_of()` doesn't check for missing variables.

```{r}
data %>% 
  dplyr::select(any_of(my_selection)) %>% 
  str()

# Won't give error message if a name is missing
data %>%
  dplyr::select(any_of(my_other_selection)) %>%
  str()

# use minus sign to remove variables
data %>%
  dplyr::select(-any_of(my_other_selection)) %>%
  str()
```

### Selection based on logical operations (`where`)

You can also select columns using functions: the `where()` selection helper selects the variables for which a function returns `TRUE`. Specifically, `where()` takes as argument a function that returns `TRUE` or `FALSE`, or purr-like formula (read the documentation to find out more). Mind that since we are dealing with logical operations, you should use the exclamation mark `!` rather than the minus `-` if you want to use `where()` to remove (= de-select) columns. You could use that to select only columns belonging to a certain type, or that have a mean above a threshold (see documentation for an example).

```{r}
# keep only columns of type character
data %>% 
  dplyr::select(where(is.character)) %>% 
  str()

# keep columns that are not character
data %>% 
  dplyr::select(!where(is.character)) %>% 
  str()

# keep columns that are character and another variable (selected by name in this instance)
data %>% 
  dplyr::select(where(is.character), mpg) %>% 
  str()

# Note that if we reverse the order of the selection, the order of the columns changes
data %>% 
  dplyr::select(mpg, where(is.character)) %>% 
  str()
```


## Extract a single column as a vector (`pull`)

While the `dplyr::select()` function takes a data frame and returns a data frame, the `pull()` function takes a data frame and returns a vector. This makes it quite similar to the dollar operator `$`. You can also use the dollar operator in pipes, relying on the dot `.` as placeholder. Remember, in R, the same thing can be achieved in many different ways.

```{r}
# 4 different ways to get the same vector from a data.frame

data %>%
  pull(mpg)

data$mpg

data %>%
  .$mpg

data %>% 
  dplyr::select(mpg) %>%
  unlist() %>%
  as.numeric()
```


## Create new variables (`mutate`)

The mutate function will be one of your best friends. To be brief, it creates new variables (i.e. columns). To be less brief, it takes a data frame, and returns a data frame with columns that are transformed in some way.
Note that the examples below are concerned with creating new variables operating on the columns, to create new variables operating on rows (such as creating summaries per group such as the mean or the number of observations, among others), you might want to look into `group_by`, `summarise` (`summarize` is a synonym), `rowwise`, etc.

```{r}
# Sum numerical variables
data %>%
  mutate(pointless_sum = mpg + cyl + disp + carb) %>%
  print()

# Don't be shy to put transformations in the sum!
data %>%
  mutate(even_more_pointless_sum = mpg + log10(cyl) + sqrt(disp)) %>%
  print()

# Apply a function, such as a logarithm
data %>%
  mutate(mpg_log10 = log10(disp)) %>%
  print()

# Change the type of variable
data %>%
  mutate(mpg = as.character(mpg)) %>%
  str()

# Add row numbers
data %>%
  mutate(row.n = row_number()) %>%
  str()
```

### Mutate multiple columns at the same time (`across`)

But what if we wanted to apply a function to many columns at once? `across(.cols = everything(), .fns)` to the rescue! You can combine with the helpers for tidy selection you have seen in the previous section.

```{r}
# apply a function to all columns (default behavior)
data %>%
  mutate(across(.fns = as.character)) %>%
  str()

# apply a function to the columns whose name meets a condition
data %>%
  mutate(across(.cols = starts_with("c"), .fns = as.character)) %>%
  str()

# apply a function to columns selected by name
data %>%
  mutate(across(.cols = c(mpg, cyl, disp), .fns = sqrt)) %>%
  str()

# write a function yourself
data %>%
  mutate(across(.cols = !where(is.character), .fns = function(x) x/10 + 5)) %>%
  str()
data %>%
  mutate(across(.cols = !where(is.character), .fns = function(x) round(x/10 + 5, digits = 0))) %>%
  str()

# apply a function and provide a regular name for the resulting variables
data %>%
  mutate(across(.cols = c(mpg, cyl, disp), .fns = sqrt, .names = "{.col}.sqrt")) %>%
  str()
```

### Adding and removing `NA`

Sometimes we might wish to transform a certain value into `NA`. You may use `na_if(x, y)` for that, which comes with two arguments: a vector to modify (`x`) and the value to replace with `NA` (`y`).

```{r}
# check the variables cyl and carb: the number 4 has been replaced by NA
# the second argument to na_if (y) is separated by a comma, although you could avoid that by writing it as a function
data %>%
  mutate(across(.cols = starts_with("c"), .fns = na_if, y = 4)) %>%
  str()
data %>%
  mutate(across(.cols = starts_with("c"), .fns = function(whatev) na_if(x = whatev, y = 4))) %>%
  str()
```

In some other cases, we might want to replace `NA` with a certain value. You may use `na_replace(data, replace)` for that, which comes with two arguments: data frame or vector (`data`) and a list of values or a single value to replace the `NA`, depending on whether the input is a data frame or vector (`replace`).


```{r}
# note that there might be easier, more elegant, and more tidy-versy ways to calculate how many values are NA per column
data %>%
  apply(2, is.na) %>%
  apply(2, sum)

# replace those NA with 9000. Note the second argument is separated by a comma
data %>%
  mutate(across(.cols = 1:3, .fns = replace_na, 9000)) %>%
  str()
```

## Sorting a data frame (`arrange()`, `desc()`)

The `arrange()` function takes a data frame, and returns a data frame in which the rows are sorted in ascending order (from small to big) based on some column(s). To sort in descending order (from big to small), combine `arrange()` with the `desc()` function: `arrange(data, desc(column))`.

Note that `NA` are sorted to the end!

The base function `sort()` has a slightly different behavior

```{r}
# sort data frame in ascending order of miles per gallon (mpg)
data %>%
  arrange(mpg)

# sort data frame in descending order of miles per gallon (mpg)
data %>%
  arrange(desc(mpg))
```

## Change column order (`relocate`)

The `dplyr::relocate()` function takes a data frame, and returns a data frame with reordered columns. Use either `.before` or `.after` arguments to indicate the position (relative to some column) where you want to relocate your column(s). Combine `.before` or `.after` with `last_col()` to specify a relocated position relative to the last column. You can use tidy selection functions introduced in previous sections.

```{r}
data %>%
  dplyr::relocate(mpg, .after = cyl) %>%
  str()

# Note the use of last_col() and the difference between .after and .before
data %>%
  dplyr::relocate(mpg, .after = last_col()) %>%
  str()
data %>%
  dplyr::relocate(mpg, .before = last_col()) %>%
  str()

# combine this with tidy selection
data %>%
  dplyr::relocate(starts_with("c"), .after = last_col()) %>%
  str()
```



## Change column name (`rename` or `select`)

The `rename()` takes a data frame, and returns a data frame with changed column names, like this: `new_name = old_name`.

```{r}
data %>%
  rename(miles.per.gallon = mpg) %>%
  str()
```

You may use quotations to handle names that contain blank spaces, although it is not advisable to include blank spaces in variables names

```{r}
data %>%
  rename("miles per gallon" = mpg) %>%
  str()

# this won't work
data %>%
  rename(miles per gallon = mpg) %>%
  str()
```


This can be enhanced with `rename_with(.cols = everything(), .fn)` and tidy selection, and applying regular expressions for the manipulation of strings.

```{r}
# change all names to uppercase, .cols defaults to everything()
data %>%
  rename_with(.fn = toupper) %>%
  str()

# this alternative syntax returns the same result
data %>%
  rename_with(.fn = ~toupper(.)) %>%
  str()

# change a few selected names to uppercase
data %>%
  rename_with(.cols = starts_with("c"), .fn = toupper) %>%
  str()

# use regular expressions
data %>% 
  rename_with(.cols = everything(), .fn = ~ str_replace(string = ., pattern = "a", replacement = "AAA")) %>% 
  str()

# alternative syntax with the same result
data %>% 
  rename_with(.cols = everything(), .fn = str_replace, pattern = "a", replacement = "AAA") %>% 
  str()
```

The `select()` function can also be used to change the name of selected columns, similarly to `rename`, specifying `new_name = old_name`.

```{r}
data %>%
  dplyr::select(miles.per.galon = mpg) %>%
  str()

data %>%
  dplyr::select(miles.per.galon = mpg, everything()) %>%
  str()

data %>%
  dplyr::select("miles per galon" = mpg, everything()) %>%
  str()
```


# Shortcuts

For a complete list of shortcuts: Tools > Keyboard Shortcuts Help.
From the whole range of shortcuts, I highlight the following:

Shortcut|Result
---|---
Control + Enter|Run code (line or selection)
Control + Shift + r|Create and name section
Control + Shift + a|Format code (line or selection)
Control + Shift + c|Comment code (line or selection)
Control + Shift + m|Write pipe `%>%`
ts + Tabulator + Enter|Create a section with the date and hour


Here's an extra tip: when you create an object, wrap the line in parentheses to print the created object. This removes the need to use an additional line of code to `print()`the object, since objects are not printed when they are created.

```{r}
sample.df <- data.frame(Var1 = c(1, 2, 3), Var2 = c(1, 2, 3))
```

```{r}
(sample.df <- data.frame(Var1 = c(1, 2, 3), Var2 = c(1, 2, 3)))
```

# Structuring your code

I like to start my scripts indicating when the script was created, as well as adding timestamps if the code has been revised substantially. Moreover, creating numbered sections with informative titles helps a lot to give order and order to the code.

* Create time stamps (`ts + Tabulator + Enter`). This will create a section with the date and hour
* Create section (`Control + Shift + r`). Note that sections can be created by writing 5 or more #

Moreover, it can be a good idea to follow a consistent order in all scripts. For instance, start the script by loading libraries and files. In addition, you may store the libraries, and even paths to the different files, in a separate scripts, and run them at the beginning of the code through the `source()` function.
